Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-10 paper/main.tex
new class
shuffle some sections, drop old intro

\documentclass[nocopyrightspace,preprint]{sigplanconf}

@@

\usepackage[numbers]{natbib}

@@

\lstset{moredelim=[is][\underbar]{_}{_},keepspaces=true}

@@

% In software engineering and software verification, we often would like
% to have static information about the quantitative behavior of
% programs.  For example, stack and heap-space bounds are important to
% ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
% % Clock-cycle bounds are needed to guarantee the safety of real-time
% % systems [XXX wilhelm].
% Static energy usage information is critical for autonomous systems and
% has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
% Worst-case time bounds can help to create constant-time
% implementations that prevent side-channel
% attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
% bounds are used to ensure the accuracy of programs that are executed
% on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
% needed to verify cryptographic protocols~\cite{BartheGB09}.
% % XXX Mention differential privacy here?
% %
% In general, quantitative performance information at design time can
% provide useful feedback for developers.
% Static analysis of quantitative properties of imperative programs is
% an active research area and recent years have seen many innovations.
% Notable tools that have been developed include
% SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
% PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
% %
% While the these tools can derive impressive results for realistic
% software, there still exist shortcomings that hamper the application
% of existing techniques in practice:
% \itemskip
% \begin{itemize}
% \item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
%   without enabling \emph{user interaction} to manually or
%   semi-automatically derive bounds for challenging parts of the
%   program.
% \itemskipIn
% \item The computed bounds are often \emph{non-compositional} local
%   bounds (for a single (nested) loop) that are difficult to combine to
%   global whole program bounds.
% \itemskipIn
% \item Existing techniques often rely on complex external tools such as
%   abstract interpretation--based invariant
%   generation~\cite{GulwaniMC09} or translation of the program into a
%   term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
%   providing \emph{verifiable certificates} for the correctness of the
%   derived bound.
% \end{itemize}
% \itemskip
% %
% While there has been much progress in static quantitative analysis,
% Knuth correctly points out in a recent interview~\cite{KnuthInter}
% that the state-of-the-art in formal quantitative methods still falls
% short in comparison with semantic techniques.
% %
% \itemskip
% \begin{quote}
%   % Consider, for example, a computer program that sorts a list of
%   % numbers into order.
%   $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
%   we have formal definitions of semantics, and tools by which we can
%   verify that sorting is indeed always achieved. My job is to go
%   beyond correctness, to an analysis of such things as the program's
%   running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
%   describes the program's performance, and all of my colleagues agree
%   with me that the recurrence is "obviously" valid. Yet I have no
%   formal tools by which I can prove that my recurrence is right. I
%   don't really understand my reasoning processes at all!
%   \vspace{-4ex}
%   \begin{flushright}
%     -- Donald E. Knuth, 2014
%   \end{flushright}
% \end{quote}
% \itemskip
% %
% In this work, we develop a resource analysis framework for C programs
% that is based on a solid semantic foundation.  The choice of C is
% primarily motivated by \iffull{our}{the} ongoing work on the formally verified
% hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
% developed in C and is supposed to provide verified guaranties on
% timing and memory usage.  Moreover, C is a natural choice because it
% is still the most widely used language for system development; in
% particular in safety-critical embedded and real-time systems where
% resource bound analyses are often required by regulatory
% authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
% %
% \itemskip
% \begin{enumerate}
% \item We define an operational cost semantics for CompCert Clight that
%   defines the resource consumption of terminating and diverging
%   executions.  The cost is parametric in a user-definable cost metric
%   and can be negative if resources are released during an execution.
% \item We develop a quantitative Hoare logic for interactively deriving
%   resource bounds for Clight programs.  The logic is implemented and
%   proved sound in the Coq Proof Assistant with respect to the cost
%   semantics of Clight.
% \itemskipIn
% \item We describe an automatic static analysis that computes bounds
%   together with derivations in the quantitative Hoare logic.  The
%   analysis is inspired by type-based amortized resource analysis for
%   functional programs~\cite{Jost03,HoffmannAH11}.  It computes
%   derivations by generating simple linear constraints that can be
%   solved by an off-the-shelf LP solver, and does not require any
%   fixpoint computations to obtain loop invariants.
% \itemskipIn
% \item We show with a publicly available prototype implementation, and
%   an experimental evaluation, that our novel amortized resource
%   analysis works precisely and efficiently for challenging loop and
%   recursion patterns, and that the derived constant factors in the
%   bounds are close to or identical with the optimal ones.
% % \item We prove the completeness of the program logic
% %   with respect to the operational cost semantics.
% \end{enumerate}
% \itemskip
% %
% To the best of our knowledge, this article presents the first resource
% analysis framework for C that makes it possible to combine non-trivial
% automatically derived bounds with interactively derived bounds in a
% proof system that produces verifiable certificates for the bounds.
% Our approach complements existing work since it provides a semantic
% foundation for the computation of bounds while still supporting
% automation that sometimes goes beyond the capabilities of existing
% techniques.  Furthermore, the automatic amortized analysis and the
% quantitative logic are naturally compositional and describe the
% resource behavior of code fragments without referring to the source
% code.  As a result, we directly derive global bounds that are
% functions of the input parameters of the program.  Another unique
% feature of our system is, that it can handle resources like memory
% that may become available during execution.
% Our starting point is a recent work~\cite{veristack14} \iffull{in
%   which we have}{that has} formally verified bounds on the stack usage
% of C programs.  A key part of this stack-bound verification is based
% on a quantitative Hoare logic for deriving abstract \emph{stack
%   bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
%   depth of (recursive) function calls.}  % For
% % non-recursive programs it is possible to compute derivations for such
% % stack bounds in the quantitative Hoare logic automatically.
% % By compiling these bounds together with the C program, we have derived
% % verified stack bounds for x86 assembly code.
% %
% Our \emph{first contribution} is a generalization of the quantitative
% Hoare logic for CompCert Clight that enables us to derive resource
% bounds that are parametric in the resource of interest.  % The
% % quantitative logic is implemented in Coq and proved sound with respect
% % to a small-step operational cost semantics for Clight.
% % that counts a
% % constant resource consumption for each atomic execution step.  The
% % resource consumption at each step is parametric in a user-definable cost
% % metric and can be negative to express the release of resources (as
% % needed, e.g., to reason about memory usage).  Furthermore, a specific
% % \emph{tick} function can be inserted in the source code to further
% % customize the cost model.
% In the implementation of the logic, we use
% a shallow embedding in Coq which makes the logic very flexible.  There
% is practically no limitation on the format of the derived resource
% bounds since they can be any function that is definable in Coq.  The
% derived bounds can also be parametric in a set of cost metrics or
% specifically apply to one fixed cost metric.
% The precision and expressivity of the quantitative Hoare logic provide
% an ample foundation for reasoning about resource consumption of Clight
% programs.  % The logic is naturally compositional and the derived
% % \emph{quantitative Hoare triples} can be used to specify the resource
% % consumption of library functions or to describe resource contracts
% % between different functions.
% However, reasoning about quantitative
% properties can often be more tedious than reasoning about intentional
% properties.  Consequently, automation is inevitable to derive resource
% bounds for large code bases like CertiKOS.  Such an automation was
% easily achievable in the case of constant stack bounds for a code base
% that does not contain recursive functions~\cite{veristack14}.  In
% general however, resource bounds have to be parametric in the
% arguments of a function and depend on the number of loop iterations
% and (recursive) function calls that are performed by the function.
% For functional programs, there exist resource analysis systems---based
% on type-based amortized resource analysis~\cite{Jost03,
%   HoffmannAH12}---that can automatically derive complex polynomial
% bounds with tight constant factors.  This type-based amortized
% resource analysis has been an inspiration in the design of the
% quantitative Hoare logic and is therefore a natural candidate to
% automate the reasoning.  Type-based amortized resource analysis works
% well for functional programs with pattern matching but it has been a
% long-time open problem to apply it to C-like programs with control
% flow that depends on integer arithmetic, negative numbers, and non-linear
% control flow.
% % example, it is a
% % long-time open problem how to use type-based amortized analysis to
% % derive bounds for functions whose resource consumption depends on
% % (possibly negative) integers and non-sequential control flow as
% % introduced by \code{break} and \code{return} statements.
% % Unfortunately, these are the functions \iffull{with non-constant resource
% % consumption}{} that are most common in CertiKOS and other system
% % software.
% Our \emph{main contribution} is an automatic amortized resource
% analysis for CompCert Clight that computes derivations in the
% quantitative Hoare logic.  It is the first automatic amortized
% resource analysis that can derive bounds that depend on negative
% integers. It also handles programs with mutually recursive functions
% as well as \code{break} and \code{return} statements.  The
% potential-based approach of amortized analysis provides a single
% mechanism for analyzing programs that need special treatment in other
% techniques; including
% \itemskip
% \begin{itemize}
% \item interaction between \emph{sequential loops or function calls} through
%   size changes of variables,
% \itemskipIn
% \item \emph{nested loops} that influence each others number of iterations,
% \itemskipIn
% \item and \emph{tricky iteration patterns} as found for instance in
%   the Knuth-Morris-Pratt algorithm for string search.
% \end{itemize}
% \itemskip
% %
% The main innovation that makes the analysis practical for C is the use
% of interval sizes in potential functions instead of the sizes of
% variables.  This leads to precise bounds of programs whose resource
% consumption can be described as a function of sizes $|[x,y]| =
% \max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
% instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
%   i=i+K)} where the step-size $K>0$ is a constant.
% %
% Despite the apparent simplicity of the new analysis system, it is able
% to reproduce results from the literature (e.g,
% SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
% abstract interpretation--based methods such as disjunctive invariant
% generation~\cite{GulwaniMC09} and symbolic backward
% execution~\cite{GulwaniZ10}.  In contrast with abstract
% interpretation--based methods, our technique does not require any
% fixpoint computations to obtain loop invariants.  The mechanism we
% designed is able to leverage local assertions such as $x < y$ that we
% collect along the branching points of the program to obtain global
% resource invariants.  We achieve this by generating a linear
% constraint system that reflects the resource cost and size changes of
% integer variables in the program; a solution of the linear program
% immediately yields a resource-usage bound for the C
% program.  % We can naturally, handle
% % advanced control flow such as break, return, iteration over negative
% % integer variables, and mutually-recursive functions.
% Following the development steps of automatic amortized analysis for
% functional programs~\cite{Jost03,HoffmannH10}, we deliberately
% restrict ourselves to linear bounds in the automatic analysis (there
% are no restrictions for the manually-derived bounds in the logic).
% More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
% where $a$ and $b$ are integer variables or constants.  The reason for
% our focus on linear bounds is mainly clarity of presentation.  We
% already experimented with an extension to multivariate resource
% polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
% rules considerably more involved and should better be described
% separately.  However, we developed the linear inference system so that
% the extension to polynomial bounds shall work smoothly.
% We have evaluated the automatic analysis with system code and examples
% from the literature.  \iffull{\pref{app:cat}}{The extended version
%   of this article~\cite{anon_extended}}{} contains more than $30$
% challenging loop and recursion patterns that we collected from open
% source software and the literature.  Our analysis can find
% asymptotically tight bounds for all but 1 of these patterns, and in
% most cases the derived constant factors are tight.  To compare our
% automatic analyzer with existing techniques, we tested our examples
% with tools such as KoAT~\cite{BrockschmidtEFFG14},
% Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
% show that the bounds that we derive are often more precise than those
% derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
% uses amortization techniques, is able to achieve a similar precision.
% %
% Several micro benchmarks demonstrate the practicality and
% expressiveness of the quantitative Hoare logic.  For example, we
% derive a logarithmic bound for a binary search function, a bound that
% depends on the contents of an array to describe the exact cost of a
% function that finds a maximal element in an array, and a linear
% bound that amortizes the cost of $k$ successive increments to a binary
% counter.
% % To show the practicality of our approach we developed a case study
% % with different real-time schedulers implemented in C.  To perform a
% % schedulability analysis that takes into account the overhead of the
% % scheduler we derived time bounds for context switching and
% % initialization of the scheduler.  We then computed time bounds for a
% % set of given user task and took account the previously computed
% % scheduler overhead to determine whether the scheduling requirements of
% % all user tasks can be met.
% % With the intention, of making the material easily accessible we first
% % informally describe the new automatic amortized analysis
% % (\pref{sec:inform}).  We then formalize the intuition with the
% % operational cost semantics an the inference rules for the amortized
% % analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
% % quantitative Hoare logic and its soundness proof that is formalized in
% % Coq.  In \pref{sec:inter} we prove the soundness of the automatic
% % amortized analysis by showing that derivation using the inference
% % rules can be seen as derivations in the quantitative logic.  We also
% % show how to integrate hand-derived bounds with automatically-derived
% % bounds.  \pref{sec:exper} contains the results of the experimental
% % evaluation and \pref{sec:related} describes related research.

@@

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1) {
    _assert(na > 0);_
    a[x]=0; _na--;_
    tick(1);
    x++; }
  if (x < N) {
    a[x]=1; _na++;_
    tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h_,lg_) {
  if (h-l > 1) {
    m = l + (h-l)/2;
    _assert(lg > 0);_
    _lg--;_ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h_,lg_);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The bound derived automatically is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

The systematic technique consists in the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched
by extra variables $\vec x$ called logical variables to form a
program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
for the logical variables is shown to be such that:
$$
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n', n' \le n \land
  (\sigma, P) \Downarrow_{n'} \sigma'.
$$
Since the annotated program and the original one are often
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

This idea is illustrated by the program on \pref{fig:xmplincaux}.
Parts of the code underlined are annotations that were added
to the original program text.  The top-level loop increments
a binary counter $k$ times.  A naive analysis of
the algorithm yields the quadratic bound $k \cdot N$.
However, the algorithm is in fact linear and its cost is bounded
by $2k + \#_1(a)$ where $\#_1(a)$ denotes
the number of one entries in the array $a$.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable $na$ is introduced.
This logical variable is a reification, in the program, of the
number $\#_1(a)$.  For example, on line 6 of the example
we are setting \lstinline{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(a)$, the logical variable $na$ is decremented.
Similarly, on line 10, an array entry which was 0 becomes 1, so
$na$ is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra assertion
\lstinline{na > 0} on line 5 is always true.  It
is done by proving inductively that $na = \#_1(a)$
and remarking that since \lstinline{a[x] == 1} is true, we must have
$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
% test is in fact semantically equivalent to \lstinline{true}.

Another simple example is given on \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are underlined in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as invariant.  This trick allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).

@@

>>>>

\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}

@@

@@

\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}

@@

@@

@@

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1) {
    @assert(na > 0);@
    a[x]=0; @na--;@
    tick(1);
    x++; }
  if (x < N) {
    a[x]=1; @na++;@
    tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h@,lg@) {
  if (h-l > 1) {
    @assert(lg > 0);@
    m = l + (h-l)/2;
    @lg--;@ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h@,lg@);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The bound derived automatically is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

The systematic technique consists in the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched
by extra variables $\vec x$ called logical variables to form a
program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
for the logical variables is shown to be such that:
$$
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n' {\le} n.\,
    (\sigma, P) \Downarrow_{n'} \sigma'.
$$
Since the annotated program and the original one are often
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

This idea is illustrated by the program on \pref{fig:xmplincaux}.
Parts of the code in blue are annotations that were added
to the original program text.  The top-level loop increments
a binary counter $k$ times.  A naive analysis of
the algorithm yields the quadratic bound $k \cdot N$.
However, the algorithm is in fact linear and its cost is bounded
by $2k + \#_1(a)$ where $\#_1(a)$ denotes
the number of one entries in the array $a$.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable $na$ is introduced.
This logical variable is a reification, in the program, of the
number $\#_1(a)$.  For example, on line 6 of the example
we are setting \lstinline{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(a)$, the logical variable $na$ is decremented.
Similarly, on line 10, an array entry which was 0 becomes 1, so
$na$ is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra assertion
\lstinline{na > 0} on line 5 is always true.  It
is done by proving inductively that $na = \#_1(a)$
and remarking that since \lstinline{a[x] == 1} is true, we must have
$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails. %  So the
% test is in fact semantically equivalent to \lstinline{true}.

Another simple example is given on \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are in blue in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as invariant.  This trick allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
